The non-random assembly of functional motifs in  \nplant-pollinator networks

### 2022-03-31 Manuscript/Main_text/Main_text_Functional_Ecology.Rmd
Updates on main text

Ecological processes leave distinct structural imprints on the species interactions that shape the topology of mutualistic networks. Detecting those relationships is not trivial since they go beyond pairwise interactions, but may get blurred when considering full network descriptors. Recent work has shown the network meso-scale can capture this important information. The meso-scale describes network subgraphs representing patterns of interactions between a small number of species (i.e., motifs) that constitute the building blocks of the whole network. Despite the possible implications of network motifs to better capture species interactions, they remain overlooked in natural plant-pollinator networks. By exploring 60 empirical plant-pollinator networks from 18 different studies with wide geographical coverage we show that some motifs are consistently under- or over-represented worldwide, suggesting that the building blocks of plant-pollinator networks are not random. Furthermore, we find that distinct motif positions describing species ecological roles (e.g., generalization and number of indirect interactions) are occupied by different plant and floral visitor groups on both trophic levels. Bees appear consistently in generalised motif positions with low number of indirect interactions while the rest of floral visitor groups are associated with more specialised motif positions with higher number of indirect interactions. All plant groups are over-represented on specialised motif positions but only tall plant species with separated sexes (e.g., woody dioecious or monoecious species) are more frequent on motif positions with low number of indirect interactions. Interestingly, the realized combinations of different species groups within a motif can not be retrieved from their joint probability distributions, indicating that group combinations are not random either. Our result highlights the non-random structure of the meso-scale on plant-pollinators networks and the association of different plant and floral visitor groups with certain motifs that involve different ecological roles at a macro-ecological scale.

@@

Traditionally, plant-pollinator research has focused on direct interactions but overlooked indirect interactions (i.e., the mediated effect between two species by a third species), such as facilitative or competitive interactions between plants for pollinators [@moeller2004; @sargent2008; @carvalheiro2014]. Despite the widespread nature of indirect interactions in ecological communities [@strauss1991], plant-pollinator research often fails to finely capture those indirect interactions with the conventional analytical tools that condense the information either by species (e.g., degree) or in global topological indices (e.g., nestedness). Nonetheless, the emerging framework of network motifs (small sub-networks observed within a given network) in plant-pollinator research allows to consider both direct and indirect interactions [@simmons2019]. Motifs are often referred to as the building blocks of a network that depict subsets of interactions [@milo2002]. The analysis of network motifs in plant-pollinator networks have revealed that the different ecological processes that govern species interactions (e.g., species abundances versus trait-matching) can lead to different patterns of indirect interactions [@simmons2020]. Yet, the global patterns of indirect interactions in real plant-pollinator networks are still unknown (e.g., over- and under- represented motifs).

By linking the structural properties of the meso-scale with the species' ecology we can advance in our understanding on the determinants of species interactions. For instance, different motifs can have different ecological meanings [@simmons2019] and the position within a motif can determine the species' functional role [@stouffer2012; @baker2015]. Yet, our understanding of the ecological implications of network motifs in real plant-pollinator networks remains in its infancy. Moreover, it is unclear how the species ecology and life-history traits determines its functional role within the network of interactions [@coux2016]. For example, large pollinators can forage larger distances [@greenleaf2007], deposit greater pollen quantities [@foldesi2021] and handle complex zygomorphic flowers in comparison with small pollinators that are restricted to lower floral complexity [@gong2009]. How this different pollinator behavior translates into their interaction topology is unknown. Similarly, recent empirical findings indicate that the meso-scale is the best descriptor of plant reproductive success [@allen2021], but little is known about how plants reproductive strategies shape their position within the network of interactions. Although some studies have evaluated plant reproductive strategies in plant-pollinator networks [@tur2013; @lazaro2020], this is an often overlooked aspect in a community context [@devaux2014] and rarely incorporated into plant-pollinator network studies. Hence, exploring how the main plant and floral visitor life-history strategies integrate with the emergent motif framework could help to progress knowledge on plant-pollinator interactions.

@@

To investigate the structure of the meso-scale in plant-pollinator networks, we gathered published plant-pollinator networks with worldwide distribution. Then, we classified the plant and floral visitor species from these networks into groups that summarize their main life-history strategies. After, we calculated the normalized motif frequencies and investigated if these frequencies are similar or not to the ones expected by chance with the help of null models. Then, for each group of plants and floral visitors we also calculated the normalized group frequency in the different motif positions and investigated their overall representation by comparing these observed frequencies with their simulated counterparts. Finally, we compared if there are differences in the observed frequencies of motif combinations in empirical networks and the expected frequencies of motif combinations calculated as the product of the probabilities of the different plant and floral visitor groups within the motif.

@@

First, plant species were grouped through hierarchical clustering into the optimal number of functional groups that summarized the main plant reproductive strategies. For this, we used the trait dataset collated in @lanuza2021 that comprised a total of 1506 species including the 503 species considered in this study (**Table S2**). This dataset consisted on 8 floral, 4 reproductive and 3 vegetative traits excluding traits with high percentage of missing values (over 30$\%$; **Table S3**). We opted to calculate the plant functional groups on this larger set of species because of the higher accuracy when delimiting functional groups with that many variables and species [@dolnicar2014]. To feed the clustering analysis, we calculated the distance between the different qualitative and scaled quantitative variables with Gower distance [@gower1971]. For this, we used the function _gowdis_ with method _ward.D2_ from the package _FD_ version _1.0-12_ [@laliberte2014]. Finally, we conducted hierarchical clustering with the function _hclust_ from the _R_ stats package version _4.0.5_ and calculated the optimal number of clusters with the function _kgs_ from the package _maptree_ version _1.4-7_ [@white2009].

Second, floral visitors were aggregated into taxonomic groups based on taxonomic rank as done similarly in other plant-pollinator studies [@fenster2004; @ollerton2009]. We opted to divide floral visitors on the taxonomic rank level and not with functional traits because the main orders of floral visitors differed in form and behaviour and had lower superior taxonomical complexity (i.e., floral visitors had 6 orders versus plants that had 38). Thus, this allowed us to group floral visitors into groups that represented approximately the main life strategies of the possible pollinators: (i) bees (Hymenoptera-Anthophila), (ii) non-bee Hymenoptera (Hymenoptera-non-Anthophila), (iii) syrphids (Syrphidae-Diptera), (iv) non-syrphids-Diptera, (v) Lepidoptera and (vi) Coleoptera. However, a minor set of species belonged to other taxonomic groups that were considered in analyses but not discussed further because of their low representation in the full set of networks (3.55$\%$ of the total interactions recorded). These taxonomic groups were 'lizards', 'birds' and 'other insects'. This last group was formed by a mix of uncommon insect taxa on the full set of networks.

@@

We calculated the frequency of all motifs up to five nodes (17 different motifs in total; see **Figure 1**) for each empirical network, by using the _bmotif_ package version _2.0.2_ [@bmotifsimmons2019]. We focused exclusively on motifs up to five nodes in our analyses for computational reasons and for the ease of visualization and interpretation of a reduced set of less complex network motifs. Throughout the entire manuscript, floral visitors occupy the top nodes of the different motifs and plants the bottom ones. To control for variation in network size, motif frequencies were normalised as a proportion of the total number of motifs within each motif class (i.e., the number of nodes a motif contains). In addition, we excluded two-node motifs (or links) from our analyses because their normalised frequencies would always equal one. Note that the aim is not comparing different networks, which may differ in sampling effort, but correctly describing the structure of each individual network. Like @simmons2020, we classified the different motifs by their average path length (mean number of links between all pairs of nodes) into four different groups: (i) complete or strong ($\overline{x} = 1.38$), (ii) fan or medium-strong ($\overline{x} = 1.48$), (iii) assymmetric complete or medium-weak ($\overline{x} = 1.60$) and (iv) core-peripheral or weak ($\overline{x} = 1.85$). In general terms, plant and floral visitor groups on more densely connected motifs (e.g., complete or fan) will experience lower number of indirect interactions in contrast to groups on less densely connected ones (e.g., medium-weak or weak).

@@

To understand if the different plant and floral visitor groups are associated with motif positions that represent different network roles, we characterized the roles of each motif positions for plant and floral visitor groups in terms of their specialization and number of indirect interactions. To that end, we calculated the specificity of each motif position following @poisot2015 as $s_i = L - l_i / L - 1$, where $L$ is the total number of groups and $l_i$ the number of interaction partners of the group $i$ in a given motif. Values of the vector $si$, range between 1, complete specialization (1 partner within the motif) and 0 complete generalism (all possible partners within the motif). Then, because all nodes within a motif are connected, we calculated the number of indirect interactions ($n_i$) for each motif position and trophic guild (i.e., plants or floral visitors) as $n_i=z_i-1$, where $z_i$ is the total number of nodes for a trophic guild in a given motif. Finally, we analysed in a single model how specialization and the number of indirect interactions were associated with the observed percentiles for each trophic guild and motif position (see paragraph above). For this, we fitted a LMM model with the help again of the _lme4_ package where we considered specialization and number of indirect interactions as response variables with the interaction by plant and floral visitor group for both response variables. Further, we considered as random factor motif position nested within study system. All model diagnostics were conducted with the help of the _Dharma_ package [@hartig2017] version _0.3.3.0_ and model outputs were visualized with the _visreg_ package [@breheny2017] version 2.7.0.

@@

Finally, we studied which combinations of plant and floral visitor groups tend to appear together within the same motifs (up to five nodes). That is, which combinations are over- or under-represented. This analysis used 57 out of the 60 networks available due to computational limitations to identify all the nodes in the motifs of the three networks with the highest number of links. To do so, for each of the 53,250 possible motif combinations, we estimated the observed and the expected probability of finding those combinations in empirical networks, respectively. Then, we determined whether the observed probabilities are likely to come from the expected probabilities or not. To calculate the observed probability ($p_i^O$) of the different groups within a motif $i$ (e.g., motif 3, 'bee' + 'bee' + 'selfing herbs'), we divided the number of times that $i$ appears in our set of empirical networks ($n_i^O$) by the sum of the number of times that each possible combination appeared: $p_i^O = n_i^O / \sum_{k=1}^{53,250} n_k^O$. To estimate the expected probability of a given motif combination $i$, $p_i^E$, firstly, we calculated the probability of finding a given group $x$ in the position $\alpha$ of $i$, $p_i(x,\alpha)$. Then, by assuming the independency of $p_i(x,\alpha)$, we computed the expected probability of the combination $i$ as the product of the probability of its pairs $(x,\alpha)$, that is, $p_i^E = \prod _{(x,\alpha )\in i} p_i(x,\alpha)$. To obtain $p_i(x,\alpha)$, we proceeded as follows. First, we calculated the number of times that the group $x$ appears in the position $\alpha$ in each empirical network $\eta$, denoted as $n_i^{\eta}(x,\alpha)$. Then, by controlling for the variation at the network level, we assessed the average value of the absolute frequency with which group $x$ appears in the position $\alpha$, $n_i(x,\alpha) = E\left [ n_i^{\eta}(x,\alpha) \right ]$, and estimated $p_i(x,\alpha)$ as $p_i(x,\alpha) = n_i(x,\alpha)/\sum_{k}n_i(k,\alpha)$. To obtain $n_i(x,\alpha)$, we fitted a LMM per motif position, where the response variable was the number of times that a given position was observed per network, the explanatory variable was the group identifier, and the random intercept was given by network identifiers nested within the study identifiers.

@@

For both floral visitor and plant groups we found over- and under-representation on different motif positions that implied different ecological roles. First, for floral visitors, we found that bees were over-represented on motif positions that involved generalism (i.e., lower specificity values) while the rest of taxonomic groups (non-syrphid Diptera, syrphids, Lepidoptera, Coleoptera and non-bee Hymenoptera) showed the opposite trend with over-representation on motif positions that implied a greater level of specialization (**Figure S6 A**). Although plants did not have as marked trends as floral visitors, all plant but tall plants with small unisexual flowers had a slight tendency to be over-represented on motif positions that implied a greater level of specialization (**Figure S6 B**). Second, regarding the number of indirect interactions, bees were over-represented in motif positions with low number of indirect interactions (**Figure S7 A**) and the rest of floral visitor groups showed no differences across positions in comparison with simulated networks. For plants, all groups showed a tendency to be over-represented on motif positions with higher number of indirect interactions (**Figure S7 B**). When we further explore the observed frequencies by number of indirect interactions for floral visitor and plant groups, we found that both bees and tall plants with small unisexual flowers were the only groups with high frequencies on motif positions associated with low number interactions in comparison with the other floral visitor and plant groups (**Figure S8 and Figure S9**).

@@

The statistical comparison with z-scores between the observed and expected plant and floral visitor group combinations showed that we can not recover the observed combinations from its probability of occurrence. In fact, most combinations are under- or over- represented and follow a modified Gaussian distribution (Johnson's $S_U$) with 65$\%$ of under-represented group combinations, 20$\%$ of no statistical difference and 15$\%$ of under-represented ones (**Figure 5**). In addition, motifs with small node combinations (i.e., 2 and 3 nodes) only appeared as under-represented. Motifs with 4 and 5-node combinations appeared in the three statistical categories but 4 node motifs had a higher proportion in the under-represented category (10% in comparison with the 1$\%$ and the 3$\%$ of the no statistical difference and over-represented categories, respectively). Finally, 5-node motifs were in similar proportions in the no statistical difference and over-represented categories (99$\%$ and 97$\%$, respectively) but they appeared in lower proportion in the under-represented category (89$\%$). When we explored the identity on the different positions of the most probable motif combinations we found that the most common plant and floral visitor groups were also the most probable on these motif combinations (lower panel **Figure 5**).

@@

All data and code would be deposited at Driad Digital repository upon acceptance. For the review process we have added our plant dataset on google sheets and can be accessed with the help of the following link: https://docs.google.com/spreadsheets/d/1q3fFNwHb7k-kpLzxfgnAHGNUrdOiUnCRQU13NKXzTX4/edit?usp=sharing

>>>>

Ecological processes leave distinct structural imprints on the species interactions that shape the topology of mutualistic networks. Detecting those relationships is not trivial since they go beyond pairwise interactions, but may get blurred when considering full network descriptors. Recent work has shown that the network meso-scale can capture this important information. The meso-scale describes network subgraphs representing patterns of interactions between a small number of species (i.e., motifs) that constitute the building blocks of the whole network. Despite the possible implications of network motifs to better capture species interactions, they remain overlooked in natural plant-pollinator networks. By exploring 60 empirical plant-pollinator networks from 18 different studies with wide geographical coverage we show that some motifs are consistently under- or over-represented worldwide, suggesting that the building blocks of plant-pollinator networks are not random. Furthermore, we find that distinct motif positions describing species ecological roles (e.g., generalization and number of indirect interactions) are occupied by different plant and floral visitor groups on both trophic levels. Bees appear less frequently in specialised motif positions with high number of indirect interactions, while the rest of floral visitor groups are infrequent in generalized motif positions with low number of indirect interactions. All plant groups tend to be over-represented on specialised motif positions, except tall plant species with separated sexes (e.g., woody dioecious or monoecious species), which are more frequent on motif positions with low number of indirect interactions. Interestingly, the realized combinations of different species groups within a motif can not be retrieved from their joint probability distributions, indicating that group combinations are not random either. Our result highlights the non-random structure of the meso-scale on plant-pollinators networks and the association of different plant and floral visitor groups with certain motifs that involve different ecological roles at a macro-ecological scale.

@@

Traditionally, plant-pollinator research has focused on direct interactions but overlooked indirect interactions (i.e., the mediated effect between two species by a third species), such as facilitative or competitive interactions between plants for pollinators [@moeller2004; @sargent2008; @carvalheiro2014]. Despite the widespread nature of indirect interactions in ecological communities [@strauss1991], plant-pollinator research often fails to finely capture those indirect interactions with the conventional analytical tools that condense the information either by species (e.g., degree) or in global topological indices (e.g., nestedness). Nonetheless, the emerging framework of network motifs [small sub-networks observed within a given network which are often referred as its building blocks\; @milo2002] allows to consider both direct and indirect interactions [@simmons2019]. The analysis of network motifs in plant-pollinator networks have revealed that the different ecological processes that govern species interactions (e.g., species abundances versus trait-matching) can lead to different patterns of indirect interactions [@simmons2020]. Yet, the global patterns of indirect interactions in real plant-pollinator networks are still unknown (e.g., over- and under- represented motifs).

By linking the structural properties of the meso-scale with the species' ecology we can advance in our understanding on the determinants of species interactions. For instance, different motifs can have different ecological meanings [@simmons2019] and the position within a motif can determine the species' functional role [@stouffer2012; @baker2015]. Yet, our understanding of the ecological implications of network motifs in real plant-pollinator networks remains in its infancy. Moreover, it is unclear how the species ecology and life-history traits determine its functional role within the network of interactions [@coux2016]. For example, large pollinators can forage larger distances [@greenleaf2007], deposit greater pollen quantities [@foldesi2021] and handle complex zygomorphic flowers in comparison with small pollinators that are restricted to lower floral complexity [@gong2009]. Despite these obvious differences in pollinator behaviour, how contrasting species' ecology translates into different interaction patterns is still unknown. Similarly, recent empirical findings indicate that the meso-scale is the best descriptor of plant reproductive success [@allen2021], but little is known about how plants reproductive strategies shape their position within the network of interactions. Although some studies have evaluated plant reproductive strategies in plant-pollinator networks [@tur2013; @lazaro2020], this is an often overlooked aspect in a community context [@devaux2014] and rarely incorporated into plant-pollinator network studies. Hence, exploring how the main plant and floral visitor life-history strategies integrate with the emergent framework of network motifs could help to progress knowledge on the ecological processes that shape plant-pollinator interactions.

@@

To investigate the structure of the meso-scale in plant-pollinator networks, we gathered published plant-pollinator networks with worldwide distribution. Then, we classified the different plant and floral visitor species from these networks into groups that summarise their main life-history strategies. After, we calculated the normalised motif frequencies and investigated if these frequencies are similar or not to the ones expected by chance with the help of null models. Then, for each group of plants and floral visitors we investigated the ecological implications of their overall representation on the different motif positions, by comparing, on the one hand, the observed (normalised) frequencies of each group in the different motif positions with the frequencies of their simulated counterparts, and, on the other hand, the differences among the patterns of connections of motif positions. Finally, we compared if there are differences in the observed frequencies of motif combinations from the empirical networks and the expected frequencies of motif combinations calculated as the product of the probabilities of the different plant and floral visitor groups within the motif.

@@

First, plant species were grouped through hierarchical clustering into the optimal number of functional groups that summarized the main plant reproductive strategies. For this, we used the trait dataset collated in @lanuza2021 that comprised a total of 1506 species including the 503 species considered in this study (**Table S2**). This dataset consisted on 8 floral, 4 reproductive and 3 vegetative traits excluding traits with high percentage of missing values (over 30$\%$; **Table S3**). We opted to calculate the plant functional groups on this larger set of species because of the higher accuracy when delimiting functional groups with that many variables and species [@dolnicar2014]. To feed the clustering analysis, we calculated the distance between the different qualitative and scaled quantitative variables with Gower distance [@gower1971]. For this, we used the function _gowdis_ with method _ward.D2_ from the _FD_ package version _1.0-12_ [@laliberte2014]. Finally, we conducted hierarchical clustering with the function _hclust_ from the _R_ stats package version _4.0.5_ and calculated the optimal number of clusters with the function _kgs_ from the _maptree_ package version _1.4-7_ [@white2009].

Second, floral visitors were aggregated into taxonomic groups based on taxonomic rank as done similarly in other plant-pollinator studies [@fenster2004; @ollerton2009]. We opted to divide floral visitors on the taxonomic rank level and not with functional traits because the main orders of floral visitors differed in form and behaviour and had lower superior taxonomic complexity (i.e., floral visitors had 6 orders versus plants that had 38). Thus, this allowed us to group floral visitors into groups that represented approximately the main life strategies of the possible pollinators: (i) bees (Hymenoptera-Anthophila), (ii) non-bee Hymenoptera (Hymenoptera-non-Anthophila), (iii) syrphids (Syrphidae-Diptera), (iv) non-syrphids-Diptera, (v) Lepidoptera and (vi) Coleoptera. However, a minor set of species belonged to other taxonomic groups that were considered in analyses but not discussed further because of their low representation in the full set of networks (3.55$\%$ of the total interactions recorded). These taxonomic groups were 'lizards', 'birds' and 'other insects'. This last group was formed by a mix of uncommon insect taxa on the full set of networks.

@@

We calculated the frequency of all motifs up to five nodes (17 different motifs in total; see **Figure 1**) for each empirical network, by using the _bmotif_ package version _2.0.2_ [@bmotifsimmons2019]. We focused exclusively on motifs up to five nodes in our analyses for computational reasons and for the ease of visualization and interpretation of a reduced set of less complex network motifs. Throughout the entire manuscript, floral visitors occupy the top nodes of the different motifs and plants the bottom ones. To control for variation in network size, motif frequencies were normalised as a proportion of the total number of motifs within each motif class (i.e., 2-species motifs, 3-species motifs, 4-species motifs and 5-species motifs). In addition, we excluded 2-species (or nodes) motifs from our analyses because its normalised frequencies would always equal one. Note that the aim is not comparing different networks, which may differ in sampling effort, but correctly describing the structure of each individual network. Like @simmons2020, we classified the different motifs by their average path length (mean number of links between all pairs of nodes) into four different groups: (i) complete or strong ($\overline{x} = 1.38$), (ii) fan or medium-strong ($\overline{x} = 1.48$), (iii) assymmetric complete or medium-weak ($\overline{x} = 1.60$) and (iv) core-peripheral or weak ($\overline{x} = 1.85$). In general terms, plant and floral visitor groups on more densely connected motifs (e.g., complete or fan) will experience lower number of indirect interactions in contrast to groups on less densely connected ones (e.g., medium-weak or weak).

@@

To understand if the different plant and floral visitor groups are associated with motif positions that represent different network roles, we characterized the roles of each motif positions for plant and floral visitor groups in terms of their specialization and number of indirect interactions. To that end, we calculated the specificity of each motif position following @poisot2015 as $s_i = L - l_i / L - 1$, where $L$ is the total number of groups and $l_i$ the number of interaction partners of the group $i$ in a given motif. Values of the vector $si$, range between 1, complete specialization (1 partner within the motif) and 0 complete generalism (all possible partners within the motif). Then, because all nodes within a motif are connected, we calculated the number of indirect interactions ($n_i$) for each motif position and trophic guild (i.e., plants or floral visitors) as $n_i=z_i-1$, where $z_i$ is the total number of nodes for a trophic guild in a given motif. Thus, we considered exclusively indirect interactions within plants though shared floral visitors or within floral visitors though shared plants. Finally, we analysed in a single model how specialization and the number of indirect interactions were associated with the observed percentiles for each trophic guild and motif position (see paragraph above). For this, we fitted a LMM model with the help again of the _lme4_ package where we considered specialization and number of indirect interactions as response variables with the interaction by plant and floral visitor group for both response variables. Further, we considered as random factor motif position nested within study system. All model diagnostics were conducted with the help of the _Dharma_ package [@hartig2017] version _0.3.3.0_ and model outputs were visualized with the _visreg_ package [@breheny2017] version 2.7.0.

@@

Finally, we studied which combinations of plant and floral visitor groups tend to appear together within the same motifs (up to five nodes). That is, which combinations are over- or under-represented. This analysis used 57 out of the 60 networks available due to computational limitations to identify all the nodes in the motifs of the three networks with the highest number of links. To do so, for each of the 53,250 possible motif combinations, we estimated the observed and the expected probability of finding those combinations in empirical networks, respectively. Then, we determined whether the observed probabilities are likely to come from the expected probabilities or not. To calculate the observed probability ($p_i^O$) of the different groups within a motif $i$ (e.g., motif 3, 'bee' + 'bee' + 'selfing herbs'), we divided the number of times that $i$ appears in our set of empirical networks ($n_i^O$) by the sum of the number of times that each possible combination appeared: $p_i^O = n_i^O / \sum_{k=1}^{53,250} n_k^O$. To estimate the expected probability ($p_i^E$) of a given motif combination $i$, we assumed that the probability of finding a given group $x$ in the position $\alpha$ is independent of the probabilities associated with other positions and groups in that motif. Therefore, the expected probability ($p_i^E$) of the motif combination $i$ is given by the product of the probabilities of its constituents ($p_i^E = \prod _{(x,\alpha )\in i} p_i(x,\alpha)$). Then, by controlling for the variation at the network level with an LMM (see below), we calculated the average number of times that the group $x$ appears in the position $\alpha$ (denoted as $n_i(x,\alpha)$). To obtain the expected probability $p_i(x,\alpha)$ we divided the number of times that the group $x$ appears in the position $\alpha$ by the the total number of times that any group was observed in that position (that is, $p_i(x,\alpha) = n_i(x,\alpha)/\sum_{k}n_i(k,\alpha)$). The average number of times that the group $x$ appears in the position $\alpha$ ($n_i(x,\alpha)$) was calculated with a LMM for each position, where the response variable was the number of times that a group was observed at a given position in each real-world network, the explanatory variable was the group identifier, and the random intercept was given by network identifiers nested within the study identifiers.

@@

For both floral visitor and plant groups we found over- and under-representation on different motif positions that implied different ecological roles. First, for floral visitors, we found that bees were less frequent on motif positions that involved specialisation (i.e., lower specificity values) while the rest of taxonomic groups (non-syrphid Diptera, syrphids, Lepidoptera, Coleoptera and non-bee Hymenoptera) were infrequent in generalized motif positions with low number of indirect interactions (**Figure S6 A**). Although plants did not show differences with null models on their frequencies, all plant but tall plants with small unisexual flowers showed a slight tendency to be over-represented on specialised motif positions (**Figure S6 B**). Second, regarding the number of indirect interactions, bees were over-represented in motif positions with low number of indirect interactions (**Figure S7 A**) and the rest of floral visitor groups showed no differences across positions in comparison with simulated networks. For plants, all groups showed a slight trend to be over-represented on motif positions with higher number of indirect interactions (**Figure S7 B**). When we further explore the observed frequencies by number of indirect interactions for floral visitor and plant groups, we found that both bees and tall plants with small unisexual flowers were the only groups with high frequencies on motif positions associated with low number interactions in comparison with the other floral visitor and plant groups (**Figure S8 and Figure S9**).

@@

The statistical comparison with z-scores between the observed and expected plant and floral visitor group combinations showed that we can not recover the observed combinations from its probability of occurrence. In fact, most combinations are under- or over- represented and follow a modified Gaussian distribution (Johnson's $S_U$) with 65$\%$ of under-represented group combinations, 20$\%$ of no statistical difference and 15$\%$ of under-represented ones (**Figure 5**). In addition, motifs with small node combinations (i.e., 2 and 3 nodes) only appeared as under-represented. Motifs with 4 and 5 node combinations appeared in the three statistical categories but 4 node motifs had a higher proportion in the under-represented category (10% in comparison with the 1$\%$ and the 3$\%$ of the no statistical difference and over-represented categories, respectively). Finally, 5 node motifs were in similar proportions in the no statistical difference and over-represented categories (99$\%$ and 97$\%$, respectively) but they appeared in lower proportion in the under-represented category (89$\%$). When we explored the identity on the different positions of the most probable motif combinations we found that the most common plant and floral visitor groups were also the most probable on these motif combinations (lower panel **Figure 5**).

@@

All data and code is deposited at GitHub: https://github.com/JoseBSL/FunctionalMotifs.
